---
title: "DataRes NBA Fouls and Referees"
author: "Kaushik Naresh"
date: "12/18/2019"
output: html_document
---

```{r setup, include=FALSE}
library("ggplot2")
knitr::opts_chunk$set(echo = TRUE)
```

#### Getting the data frames
```{r}
data <- read.csv("refrefineddata.csv") # data frame of match data, used for counting number of matches...
teamref <- read.csv("Team By Referee.csv") # data frame of total fouls per team for each referee
matchref <- read.csv("Match By Referee.csv") # data frame of fouls per game for each referee (3 refs per game)
head(teamref)
head(matchref) 
```


#### This block creates a matrix of fouls against each team by each referee, essentially contains the same data as the data frame teamref
```{r}

refnames <- unique(matchref$ref)
teams <- unique(matchref$teamAbbr)
ans_mat <- matrix(nrow = length(refnames), ncol = length(teams))
refnames <- sort(refnames)
teams <- sort(teams)
colnames(ans_mat) <- teams
rownames(ans_mat) <- refnames
for ( i in refnames){
  for (j in teams){
    ans_mat[i,j] <- sum(matchref$teamPF[matchref$ref == i & matchref$teamAbbr == j]) + sum(matchref$opptPF[matchref$ref == i & matchref$opptAbbr == j])
  }
}
```





#### Total fouls be referee and by team
```{r}
#creating a named vector with total number of fouls against each team
fouls_team <- numeric(ncol(ans_mat))
for ( i in 1:ncol(ans_mat)) fouls_team[i] <- sum(ans_mat[,i])/3 # division by three because 3 refs count each foul
names(fouls_team) <- teams

#creating a named vector with total number of fouls called by each referee
fouls_ref <- numeric(nrow(ans_mat))
for ( i in 1:nrow(ans_mat)) fouls_ref[i] <- sum(ans_mat[i,])
names(fouls_ref) <- refnames

fouls_team
fouls_ref
```

---

#### Graphs of total fouls by team and total fouls by referee
```{r}
barplot(sort(fouls_team), xlab = "Team", ylab = "Number of Fouls",  cex.names = 0.3, main= "Barplot of Fouls by Team")
```

As can be seen from this graph, all teams have had at least 8000 fouls called against them in the years 2012-2018, and the three teams with the most fouls called are the Phoenix Suns, Toronto Raptors, and Philadelphia 76ers while the three teams with the lowest fouls are the Atlanta Hawks, San Antonio Spurs, and Charlotte Hornets. However, a problem here is that some teams may have played more games than others (due to playoffs) and so we need to look at averages.

```{r}
plot(sort(fouls_ref), ylab = "Number of Fouls", main = "Graph of Fouls per Referee", xlab = "Index")
```

As can be seen from this graph, more than half of the referees of the years 2012-2018 have called more than 10000 fouls and around 10% have called more than 15000 fouls while around a quarter have called less than 5000 fouls. The problem with this graph is that some referees might have officiated more games than the others, so we need to check the fouls called per game. It is difficult to label these points, so a detailed analysis of the extremes will follow.



#### Calculating average fouls by team and average fouls by referee
```{r}
# create vector for number of matches per team & then average fouls done
matches_team <- numeric(length(fouls_team))
for (i in 1:length(matches_team)) matches_team[i] <- sum(data$teamAbbr == names(fouls_team)[i] | data$opptAbbr == names(fouls_team)[i])
avg_team <- fouls_team/matches_team

#create vector for number of matches per referee & then average fouls called
matches_ref <- numeric(length(fouls_ref))
for (i in 1:length(matches_ref)) matches_ref[i] <- sum(matchref$ref == names(fouls_ref)[i])
avg_ref <- fouls_ref/matches_ref

```

#### Graphs of average fouls by team and average fouls by referee
```{r}
barplot(sort(avg_team), xlab = "Team", ylab = "Average Fouls",  cex.names = 0.3, main= "Barplot of Average Fouls by Team")

```

As can be seen, the average number of fouls per team for all teams is definitely above 17, and much like the previous barplot, the top 3 (Phoenix Suns, Toronto Raptors, Philadelphia 76ers) and bottom 3 (Charlotte Hornets, San Antonio Spurs, Atlanta Hawks) teams are the same. However, the order of teams in the middle is slightly different in this plot. Still, there does not appear to be too much of a difference suggesting that over the 6 years, these teams have probably all played around the same number of games even after factoring playoffs.

```{r}
boxplot(avg_team, main = "Boxplot of Average Fouls Per Team", ylab = "Average Fouls")
```

This boxlpot shows that the average number of fouls called lies between 18 and 22 for all teams with the median at around 20. There are also no outliers in this plot.


```{r}
plot(sort(avg_ref), ylab = "Average Fouls Called", main = "Graph of Average Fouls per Referee", xlab = "Index")
```

From this graph we can see that most referees, around 80% call between 38 and 42 fouls per game, which comes to 19-21 fouls per team per game, suggesting that there are very few referees who call an excessive amount, or a lower amount of fouls. It is difficult to label these points, so a detailed analysis of the extremes will follow.

```{r}
boxplot(avg_ref, main = "Boxplot of Average Fouls Called by Referees", ylab = "Average Fouls Called")
```

As can be seen, this boxplot confirms the previous analysis since all the outliers lie outside of the 38-42 range and the median is at around 40.

---

Now moving on to the referees that call most fouls and least fouls. (sorted by average)
#### Creating dataframes for the referees that call the least and most fouls
```{r}
ref_least <- names(sort(avg_ref)[1:3])
ref_most <- names(sort(avg_ref, decreasing = TRUE)[1:3])
df_leastref <- teamref[teamref$ref == ref_least[1] | teamref$ref == ref_least[2] | teamref$ref == ref_least[3],]
df_mostref <- teamref[teamref$ref == ref_most[1] | teamref$ref == ref_most[2] | teamref$ref == ref_most[3],]
df_mostref <- df_mostref[df_mostref$fouls != 0,]
df_leastref <- df_leastref[df_leastref$fouls != 0,]
```


#### Graphs for the top 3 most calls and lowest 3 calls
```{r}
ggplot(data = df_mostref, aes(x = ref, y = fouls, color = team)) + geom_point() + labs(title = "Graph of Fouls Called by Top 3 Referees", x = "Referee", y = "Total Fouls")
```

These are the top 3 referees who call the highest average number of fouls. As can be seen, Scott Bolnick and Vladamir Voyard Tadal do not seem to call too many fouls for any certain team in comparison to other teams. However, Scott Foster seems to have called a lot more fouls against Dallas, Chicago, and Houston, while he called much lesser fouls against Miami, Detroit and Minnesota.

Interestingly there are articles about Scott Foster where players have expressed a dislike towards his officiating. A quote from Paul Pierce (who played for Boston, LA Clippers, Brooklyn) is as follows: "I know that Scott is going to have a quick trigger with the technicals." Now as can be seen on the graph he has called a large number of fouls against each team, around 500+ for most, and his average number of fouls is 42 per game, which is definitely a pretty high number.


```{r}
ggplot(data = df_leastref, aes(x = ref, y = fouls, color = team)) + geom_point() + labs(title = "Graph of Fouls Called by Bottom 3 Referees", x = "Referee", y = "Total Fouls")
```

As can be seen, in this graph there is a much wider range. Not counting the zeroes, Matt Myers has called much more fouls against Portland and Milwaukee than Cleveland and Minnesota. Randy Richardson has called way more fouls against Miami and Sacramento than
Portland and New York. SirAllen Conner on the other hand has called a large number of fouls against Cleveland compared to the other teams.

Matt Myers is from the Midwest and interstingly, he has called lesser fouls against multiple midwest teams (Cleveland, Minnesota, Chicago). One important fact to note is that all these referees are replacement referees and did not referee for many seasons. This could be related to why they call the lowest average number of fouls.

---

Now moving on to the team with the most fouls and least fouls. (sorted by average)
#### creating the data frames for the team with most fouls (Phoenix) and the teams with least fouls (Charlotte)
```{r}
max_team <- names(sort(avg_team, decreasing = TRUE))[1]
min_team <- names(sort(avg_team))[1]

df_teammax <- teamref[teamref$team == max_team,]
df_teammin <- teamref[teamref$team == min_team,]

df_teammax <- df_teammax[df_teammax$fouls != 0,]
df_teammin <- df_teammin[df_teammin$fouls != 0,]

x <- quantile(df_teammax$fouls, probs = c(0.1, 0.9))
y <- quantile(df_teammin$fouls, probs = c(0.1, 0.9) )

df_teammax <- df_teammax[df_teammax$fouls > x[2] | df_teammax$fouls < x[1],]
df_teammin <- df_teammin[df_teammin$fouls > y[2] | df_teammin$fouls < y[1],]
```

#### graphs
```{r}
ggplot(data=df_teammax, aes(x = team, y = fouls, color = ref)) + geom_point() + labs(title = "Graph of Fouls Called Against Phoenix Suns", x = "Team", y = "Total Fouls")

```


```{r}
ggplot(data=df_teammin, aes(x = team, y = fouls, color = ref)) + geom_point() + labs(title = "Graph of Fouls Called Against Charlotte Hornets", x = "Team", y = "Total Fouls")
```


